---
title: "ESM 244 Lab 6"
author: "Patrick Pelegri-O'Day"
date: "2/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(palmerpenguins)
library(here)

library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```

## Part 1: k-means clustering

```{r}
### bill length vs bill depth
ggplot(penguins) + 
  geom_point(aes(x = bill_length_mm, y = bill_depth_mm,
                 color = species, shape = sex),
             size = 3, alpha =0.7)
```

```{r}
ggplot(penguins) + 
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = species, shape = sex))
```


Note that the variables are not on the same scale - some are orders of magnitude larger/smaller
So let's rescale each one so that they're comparable
```{r}
summary(penguins)
```

Also note from the summary the NAs. We can drop them because there aren't too many
```{r}
penguins_complete <- penguins %>% 
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

penguins_scale <- penguins_complete %>% 
  select(ends_with(c('mm', 'g'))) %>% 
  scale()
```

### How many clusters?
```{r}
number_est <- NbClust(penguins_scale, min.nc = 2, max.nc = 10,
                      method = 'kmeans')

### knee method
fviz_nbclust(penguins_scale, FUNcluster = kmeans, method = 'wss', k.max = 10)
```

### let's run some k-means

```{r}
# nstart refers to how many iterations we run
penguins_km <- kmeans(penguins_scale, centers = 3, nstart = 25)
penguins_km$size
penguins_km$cluster
penguins_km$centers

penguins_cl <- data.frame(penguins_complete,
                          cluster_number  = factor(penguins_km$cluster))

ggplot(penguins_cl) + 
  geom_point(aes(x = flipper_length_mm, y = body_mass_g, color = cluster_number, shape = species))

# compare clustering to species
penguins_cl %>% select(species, cluster_number) %>%  table()
```

## Part 2: hierarchical cluster analysis (agglomerative)

```{r}
### create distance matrix
peng_dist <- dist(penguins_scale, method = 'euclidean')

### hierarchical clustering (complete linkage)
peng_hc_complete <- hclust(peng_dist, method = 'complete')

### plot suing base plot
plot(peng_hc_complete, cex = 0.6, hang = -1)
```

```{r}
### cut into three clusters
peng_cut_hc <- cutree(peng_hc_complete, 3)

table(peng_cut_hc, penguins_complete$species)
```

### World Bank data - read in and simple

```{r}
wb_env <- read_csv(here('data', 'wb_env.csv'))
```

```{r}
wb_ghg_20 <- wb_env %>% 
  arrange(-ghg) %>% 
  head(20) # could also use slice_map

summary(wb_ghg_20) # hugely different scales; need to re-scale!
```

### scale the data

```{r}
wb_scaled <- wb_ghg_20 %>% 
  select(3:7) %>% # we ditch the columns with characters because scale() can only scale if all numbers
  scale()

rownames(wb_scaled) <- wb_ghg_20$name
```

### Calc distance matrix

```{r}
euc_distance <- dist(wb_scaled, method = 'euclidean', diag = TRUE, upper = TRUE)
```

### do our clustering

```{r}
hc_complete <- hclust(euc_distance, method = 'complete')

plot(hc_complete, cex = 0.6, hang = -1)
```

```{r}
hc_single <- hclust(euc_distance, method = 'single')

plot(hc_single, cex = 0.6, hang = -1)
```

### make a tanglegram

```{r}
### convert to class dendrogram
dend_complete <- as.dendrogram(hc_complete)
dend_single <- as.dendrogram(hc_single)

tanglegram(dend_complete, dend_single)
```

### Make a ggplot dendrogram

```{r}
ggdendrogram(hc_complete, rotate = TRUE) +
  theme_minimal() +
  labs(x = 'Country', y = 'Distance')
```

